Statistical Inference for inter-arrival times of extreme events in bursty time series

### 2018-12-03 article_springer/article_springer.Rmd
Add plots tail parameter Exponential Example

```{r LRT power, echo=FALSE, fig.cap="\\label{Fig:LRT} Power for bootstrapped LRT for different sample sizes", out.width = '100%'}
knitr::include_graphics("Plots_LRT_boot.pdf")
```

@@

According to Figure \ref{fig:flares}, for a threshold $\ell$ at the $k$-th order 
statistic, the fitted threshold exceedance time distribution is 
$$
T(\ell) \sim {\rm ML}(\beta, k^{-1/\beta} \sigma_0), 
$$
where $\beta = 0.85$ and $\sigma_0 = 3.0 \times 10^7 {\rm sec}$. 
Unlike the exponential distribution, the Mittag-Leffler distribution is not memoryless, and the probability density of the time $t$ until the next threshold crossing will depend on the time $t_0$ elapsed since the last threshold crossing.
This density equals 
$$
p(t|\beta, \sigma_0, \ell, t_0) = \frac{f(t + t_0 | \beta, k^{-1/\beta} \sigma_0)}{\mathbf P[T_\ell > t_0]}
$$
where $f(\,\cdot\, | \beta, k^{-1/\beta} \sigma_0)$ is the probability density of ${\rm ML}(\beta, -k^{1/\beta} \sigma_0)$. 
The more time has passed without a threshold crossing, the more the probability distribution shifts towards larger values for the next crossing
(see Figure \ref{fig:hazard}, left panel). The hazard rate 
$$
h(t) = \frac{f(t| \beta, k^{-1/\beta} \sigma_0))}{\int_t^\infty f(\tau| \beta, k^{-1/\beta} \sigma_0))\,d\tau}
$$
represents the risk of a threshold crossing per unit time, and is a decreasing function for the Mittag-Leffler distribution. 
```{r hazard, fig.height=5, fig.cap="Left: Conditional distribution of time until next threshold crossing, depending on elapsed time $t_0$ since last crossing ($\\beta = 0.8$, $\\sigma_0 = 1$). Right: Hazard rate depending on tail parameter $\\beta$.\\label{fig:hazard}"}
tail <- 0.7
scale <- 1
t_0 <- c(0, 1, 10)
from <- 0.1
to <- 100 * scale
xx <- exp(seq(from = log(from), to = log(to), length.out = 100))
par(mfrow = c(1,2))
plot(xx, rep(1, length(xx)), ylim = c(0.001,1), type = 'n', log = 'xy', 
     xlab = 't', ylab = 'p')
for (k in 1:length(t_0)) {
  cond <- pml(q = t_0[k], tail = tail, scale = scale, lower.tail = FALSE)
  yy <- dml(x = xx + t_0[k], tail = tail, scale = scale) / cond
  lines(xx,yy, col = k+1, lty = k)
}
legend("bottomleft", c("t_0 = 0", "t_0 = 1", "t_0 = 10"), col = 2:4, lty = 1:3)

@@

To test our inference method, we have simulated independent waiting time and magnitude pairs $(W_k, J_k)$. The distribution of $J_k$ is irrelevant for the inference on $\beta$ and $\sigma_0$ (it just plays a role for the rescaling of $\hat{\sigma}_0$, which gets done to receive an estimate for the scale parameter of the exceedances). Hence we choosed exponential distributed magnitudes in all considered examples. For the waitings times, we choosed (i) a Stable, (ii) a Pareto and (iii) an Exponential distribution.   
In order to have exact analytical values available for $\beta$ and $\sigma_0$, a distribution for $W_k$ needs to be chosen for which $b(n)$ from \eqref{eq:stability} is known. In our first example, we choose $W_k \stackrel{d}{=} D$, where $D$ is as in \eqref{eq:stability}, then due to the stability property we have the *equality* of distribution $W_1 + \ldots + W_n \stackrel{d}{=} b(n) D$, for $b(n) = n^{1/\beta}$. Using the parametrisation of @SamorodnitskyTaqqu, a few lines of calculation (see e.g. the vignette on parametrisation in @MittagLeffleR) show that $D$ must have the stable distribution $S_\beta(\cos(\pi \beta/2)^{1/\beta}, +1, 0)$, which is implemented in the R package `stabledist` by @stabledist. If we choose $W_k$ to be Pareto distributed or more precisecly 
\begin{align*}
P(W_k>t)= C t^{-\beta}
\end{align*}
with $C:=(1/\Gamma(1-\beta))^{1/\beta}$ and $b(n)=n^{1/\beta}$ we get 
\begin{align*}
\frac{W_1 + \ldots + W_n}{b(n)} \stackrel{d}{\Rightarrow} D
\end{align*}
with $E(\exp(-sD))=\exp(-s^{\beta})$.

```{r make-caption}
caption = "Tail and scale estimates for simulated data, with waiting times drawn from the stable distribution $S_\\beta(\\cos(\\pi \\beta/2)^{1/\\beta}, +1, 0)$ with $\\beta = 0.8$. Dashed lines are 95\\% confidence intervals, dotted lines are the known theoretical values ($0.8$ and $10000^{1/0.8}$). \\label{fig:sim}"
```

```{r simulated-example, message=FALSE, fig.cap=caption}
k_max <- 500 
thin_sim_ctre <- thin(ctre = sim_ctre, k = k_max)
par(mfrow=c(1,2))
MLestimates(ctre = thin_sim_ctre, tail = params$tail, scale = n^(1/params$tail))
```

By the Theorem, the distribution 
of $T(\ell)$ is approximately 
$$
{\rm ML}(\beta, p^{-1/\beta}) 
= {\rm ML}(\beta, k^{-1/\beta} n^{1/\beta}),
$$
which means 
$\sigma_0 = n^{1/\beta}$.
The distribution of $J_k$ is irrelevant for the inference on $\beta$ and 
$\sigma_0$ (we have chosen unit exponential random variables). 
Figure \ref{fig:sim} displays plots of $\hat \beta(\ell)$ and 
$\hat \sigma(\ell)$ vs. $k$; recall that $k$ is the index of the order
statistics of $J_k$ at which the threshold $\ell$ is placed. 
Dotted lines show 95% confidence intervals, which are derived from the 
asymptotic normality of the log-moments estimators [@Cahoy2013] and the $\delta$-method [@MittagLeffleR].
The dashed lines show the actual values of $\beta$ resp.\ $\sigma_0$, 
showing that our inference method identifies the 
parameters correctly.

```{r Ex1figs, echo=FALSE,fig.height=7,fig.cap="\\label{Fig:Ex1figs} Stability Plots for m=100 simulation runs for Example 1 (stable distributed waiting times, true tail parameter =0.8). The grey lines, the red line and the blue lines are the log Moment, the MLE and the hill estimator, repectively, for different $k$'s on the x-axis. The dark lines in each picture are the means of the concerning estimators for different $k$'s.", fig.show='hold', out.width = '32%'}
knitr::include_graphics(c("SimuStudy_StabilityPlots_Tail08_logMoms.pdf","SimuStudy_StabilityPlots_Tail08_mles.pdf","SimuStudy_StabilityPlots_Tail08_hill.pdf"))
```

```{r Ex2figs, echo=FALSE,fig.height=7,fig.cap="\\label{Fig:Ex2figs} Stability Plots for m=100 simulation runs for Example 2 (Pareto distributed waiting times, true tail parameter =0.8). The grey lines, the red line and the blue lines are the log Moment, the MLE and the hill estimator, repectively, for different $k$'s on the x-axis. The dark lines in each picture are the means of the concerning estimators for different $k$'s.", fig.show='hold', out.width = '32%'}
knitr::include_graphics(c("SimuStudy_StabilityPlots_Ex2_Tail08_logMoms.pdf","SimuStudy_StabilityPlots_Ex2_Tail08_mles.pdf","SimuStudy_StabilityPlots_Ex2_Tail08_hill.pdf"))
```
# Data example

@@

Figure \ref{fig:flares} shows the stability plots for the solar flare data, on the left for the tail parameter and on the right for the scale parameter. The stability plot for the tail stabilizes nicely round about 0.85 (dotted line), while it is a little bit harder to deduce an estimate for the scale parameter. Howerever, the stability plot for the scale parameter seems to stabilize round about $3 \times 10^7$.

@@

The fit with a Mittag-Leffler distribution ($\beta = 0.85$) is good, though there are signs that the power-law tail tapers off for very large inter-threshold crossing times. There is no apparent dependence between threshold exceedance times and event magnitudes seen in the copula plot. We also counduct a bootstrapped LRT for the null hypothesis of exponential distributed inter-arrival times and received a $p$-value of

>>>>

```{r LRT power, echo=FALSE, fig.cap="\\label{Fig:LRT} Power for bootstrapped LRT for different sample sizes", out.width = '90%'}
knitr::include_graphics("Plots_LRT_boot.pdf")
```

@@

According to Figure \ref{fig:flares}, for a threshold $\ell$ at the $k$-th order 
statistic, the fitted threshold exceedance time distribution is 
$$
T(\ell) \sim {\rm ML}(\beta, k^{-1/\beta} \sigma_0), 
$$
where $\beta = 0.85$ and $\sigma_0 = 3.0 \times 10^7 {\rm sec}$. 
Unlike the exponential distribution, the Mittag-Leffler distribution is not memoryless, and the probability density of the time $t$ until the next threshold crossing will depend on the time $t_0$ elapsed since the last threshold crossing.
This density equals 
$$
p(t|\beta, \sigma_0, \ell, t_0) = \frac{f(t + t_0 | \beta, k^{-1/\beta} \sigma_0)}{\mathbf P[T_\ell > t_0]}
$$
where $f(\,\cdot\, | \beta, k^{-1/\beta} \sigma_0)$ is the probability density of ${\rm ML}(\beta, -k^{1/\beta} \sigma_0)$. 
The more time has passed without a threshold crossing, the more the probability distribution shifts towards larger values for the next crossing
(see Figure \ref{fig:hazard}, left panel). The hazard rate 
$$
h(t) = \frac{f(t| \beta, k^{-1/\beta} \sigma_0))}{\int_t^\infty f(\tau| \beta, k^{-1/\beta} \sigma_0))\,d\tau}
$$
represents the risk of a threshold crossing per unit time, and is a decreasing function for the Mittag-Leffler distribution. 
```{r hazard, fig.height=4, fig.cap="Left: Conditional distribution of time until next threshold crossing, depending on elapsed time $t_0$ since last crossing ($\\beta = 0.8$, $\\sigma_0 = 1$). Right: Hazard rate depending on tail parameter $\\beta$.\\label{fig:hazard}"}
tail <- 0.7
scale <- 1
t_0 <- c(0, 1, 10)
from <- 0.1
to <- 100 * scale
xx <- exp(seq(from = log(from), to = log(to), length.out = 100))
par(mfrow = c(1,2))
plot(xx, rep(1, length(xx)), ylim = c(0.001,1), type = 'n', log = 'xy', 
     xlab = 't', ylab = 'p')
for (k in 1:length(t_0)) {
  cond <- pml(q = t_0[k], tail = tail, scale = scale, lower.tail = FALSE)
  yy <- dml(x = xx + t_0[k], tail = tail, scale = scale) / cond
  lines(xx,yy, col = k+1, lty = k)
}
legend("bottomleft", c("t_0 = 0", "t_0 = 1", "t_0 = 10"), col = 2:4, lty = 1:3)

@@

To test our inference method, we performed a simulation study where we have simulated independent waiting time and magnitude pairs $(W_k, J_k)$. Since the estimation of the parameters of the inter-exceedance times is based on stability plots and is hence rather a graphical estimation procedure, it is difficult to conduct a simulation study. To get an impression of how our estimation procedure performs, we plotted $m=100$ stability plots for each example and calculated for each $k$ the mean value of all estimates. We have chosen (i) a Stable, (ii) a Pareto and (iii) an Exponential distribution for waiting times.  The distribution of $J_k$ is irrelevant for the inference on $\beta$ and $\sigma_0$ (it just plays a role for the rescaling of $\hat{\sigma}_0$, which gets done to receive an estimate for the scale parameter of the exceedances). Hence we have chosen exponential distributed magnitudes in all considered examples.  
In order to have exact analytical values available for $\beta$ and $\sigma_0$, a distribution for $W_k$ needs to be chosen for which $b(n)$ from \eqref{eq:stability} is known. In our first example, we have chosen $W_k \stackrel{d}{=} D$, where $D$ is as in \eqref{eq:stability}. Then due to the stability property we have the *equality* of distribution $W_1 + \ldots + W_n \stackrel{d}{=} b(n) D$, for $b(n) = n^{1/\beta}$. Using the parametrisation of @SamorodnitskyTaqqu, a few lines of calculation (see e.g. the vignette on parametrisation in @MittagLeffleR) show that $D$ must have the stable distribution $S_\beta(\cos(\pi \beta/2)^{1/\beta}, +1, 0)$, which is implemented in the R package `stabledist` by @stabledist. In the second examaple we have chosen $W_k$ to be Pareto distributed or more precisecly 
\begin{align*}
P(W_k>t)= C t^{-\beta}
\end{align*}
with $C:=(1/\Gamma(1-\beta))^{1/\beta}$ and $b(n)=n^{1/\beta}$. We then receive
\begin{align*}
\frac{W_1 + \ldots + W_n}{b(n)} \stackrel{d}{\Rightarrow} D
\end{align*}
with $E(\exp(-sD))=\exp(-s^{\beta})$.

By the Theorem, the distribution 
of $T(\ell)$ is approximately 
$$
{\rm ML}(\beta, p^{-1/\beta}) 
= {\rm ML}(\beta, k^{-1/\beta} n^{1/\beta}),
$$
which means 
$\sigma_0 = n^{1/\beta}$. we have chosen $\beta=1$ in the stable as well in the Pareto case. Figure \ref{Ex1figs} displays plots of $\hat \beta(k)$ vs. $k$ for the first example with stable distributed waiting times. In the left picture we used the log moment estimator, in the middle picture maximum likelihood estimates (both applied to the inter-exceedance times) and for the right picture hill estimators applied to the inter-arrival times. Recall that $k$ is the index of the order statistics of $J_k$ at which the threshold $\ell$ is placed.

```{r Ex1figs, echo=FALSE,fig.height=7,fig.cap="\\label{Fig:Ex1figs} Stability Plots for m=100 simulation runs for Example 1 (stable distributed waiting times, true tail parameter =0.8). The grey lines, the red line and the blue lines are the log Moment, the MLE and the hill estimator, repectively, for different $k$'s on the x-axis. The dark lines in each picture are the means of the concerning estimators for different $k$'s.", fig.show='hold', out.width = '32%'}
knitr::include_graphics(c("SimuStudy_StabilityPlots_Tail08_logMoms.pdf","SimuStudy_StabilityPlots_Tail08_mles.pdf","SimuStudy_StabilityPlots_Tail08_hill.pdf"))
```
Figure \ref{Ex2figs} shows the estimates fordifferent $k$'s and for the different estimators for the tail parameter $\beta$. Again one can see, that there is much more variance between the stability plots for the different simulation runs in case of the hill estimator compared to the other two estimators.  
```{r Ex2figs, echo=FALSE,fig.height=7,fig.cap="\\label{Fig:Ex2figs} Stability Plots for m=100 simulation runs for Example 2 (Pareto distributed waiting times, true tail parameter =0.8). The grey lines, the red line and the blue lines are the log Moment, the MLE and the hill estimator, repectively, for different $k$'s on the x-axis. The dark lines in each picture are the means of the concerning estimators for different $k$'s.", fig.show='hold', out.width = '32%'}
knitr::include_graphics(c("SimuStudy_StabilityPlots_Ex2_Tail08_logMoms.pdf","SimuStudy_StabilityPlots_Ex2_Tail08_mles.pdf","SimuStudy_StabilityPlots_Ex2_Tail08_hill.pdf"))
```
To show how the different estimates behave in the case that $\beta=1$, hence, the inter-exceedance times are exponetial distributed, we simulated in a last example exponetial distributed waiting times.

```{r Ex3figs, echo=FALSE,fig.height=7,fig.cap="\\label{Fig:Ex3figs} Stability Plots for m=100 simulation runs for Example 3 (Exponentially distributed waiting times, true tail parameter =0.8). The grey lines, the red line and the blue lines are the log Moment, the MLE and the hill estimator, repectively, for different $k$'s on the x-axis. The dark lines in each picture are the means of the concerning estimators for different $k$'s.", fig.show='hold', out.width = '32%'}
knitr::include_graphics(c("SimuStudy_StabilityPlots_Ex3_logMoms.pdf","SimuStudy_StabilityPlots_Ex3_mles.pdf","SimuStudy_StabilityPlots_Ex3_hill_times.pdf"))
```

# Data example

@@

Figure \ref{fig:flares} shows the stability plots for the solar flare data, on the left for the tail parameter and on the right for the scale parameter. Dotted lines show 95% confidence intervals, which are derived from the asymptotic normality of the log-moments estimators [@Cahoy2013] and the $\delta$-method [@MittagLeffleR, dashed lines show the actual values of $\beta$ resp.\ $\sigma_0$. The stability plot for the tail stabilizes nicely round about 0.85 (dotted line), while it is a little bit harder to deduce an estimate for the scale parameter. Howerever, the stability plot for the scale parameter seems to stabilize round about $3 \times 10^7$.

@@

The fit with a Mittag-Leffler distribution ($\beta = 0.85$) is good, though there are signs that the power-law tail tapers off for very large inter-threshold crossing times. There is no apparent dependence between threshold exceedance times and event magnitudes seen in the copula plot. We also counduct a bootstrapped LRT for the null hypothesis of exponential distributed inter-arrival times and received a $p$-value of $p<0.01$.
